---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval=FALSE,
  warning = FALSE,
  message = FALSE
)
```


# scib

<!-- badges: start -->
<!-- badges: end -->

The goal of scib (single-cell integration and benchmarking) is to provide a unified way to perform data integration and benchmarking when the ground truth is available.


# INSTALLATION 

Step 1: Install the R package dependencies:

```{r, eval=FALSE}
if(!require(pacman)) install.packages("pacman")

if(!require(RcppPlanc)) install.packages('RcppPlanc', 
                                         repos = 'https://welch-lab.r-universe.dev')

pacman::p_load(tidyverse, Seurat, SeuratWrappers,
               remotes, reticulate, leidenAlg,
               conos, harmony, rliger)

if(!require(scMC)) install_github("amsszlh/scMC")

if(!require(scib)) install_github("mthimma/scib")
```

Step 2: Install the miniconda environment (if it does not exist)

```{r, eval=FALSE}
install_miniconda()
```

Step 3: Create the conda environment

```{r, eval=FALSE}
system('curl -LJO https://raw.githubusercontent.com/mthimma/scib/refs/heads/main/scib.yaml')

conda_create(envname = "scib", environment = "scib.yaml")
```


Later ...

```{r}
python_path <- conda_list() %>% 
  filter(name=="scib") %>% 
  pull(python)

use_python(python_path)

conda_env <- dirname(python_path)  ## needed inside IntegrateLayers2() function
```


py_module_available("leidenalg")
py_module_available("pandas")


## Install Dependencies for Windows OS

Install miniforge

https://github.com/conda-forge/miniforge/releases/latest


This installs "Miniforge prompt". Open this terminal


reticulate R package




## Install Dependencies for Windows OS (old)

1. Install the R package `reticulate`

```{r, eval=FALSE}
install.packages("reticulate")
```

where.exe python
C:\Users\aramasamy\AppData\Local\mambaforge\python.exe


use_python("C:\\Users\\aramasamy\\AppData\\Local\\mambaforge\\envs\\scib\\python.exe")


2. Load the R package

```{r, eval=FALSE}
library(reticulate)
```

3. Create a conda environment via reticulate.

```{r, eval=FALSE}
conda_create(envname = "scib", python_version = "=3.11")
```

4. Install dependencies: `python-annoy` (needed for bbknn and scanorama) and `pandas`, `leidenalg` (needed for scMC).

```{r, eval=FALSE}
py_install(envname  = "scib", packages = "python-annoy", channel = "conda-forge")

py_install("python-igraph")
py_install("leidenalg", forge = TRUE)
py_install(envname  = "scib", packages = c("leidenalg", "panda"))
```

5. Install the data integration softwares.

```{r, eval=FALSE}
conda_install(envname = "scib", 
              packages = c("scvi-tools", "scanpy", "scpython", "anndata", "bbknn", "scanorama"),
              pip = TRUE)
```

6. Record the path to the `scib` environment.

```{r, eval=FALSE}
conda_list()
```


## Install Dependencies for Linux OS

1. Install miniforge conda if not available.

```{bash, eval=FALSE}
curl -L -O "https://github.com/conda-forge/miniforge/releases/latest/download/Miniforge3-$(uname)-$(uname -m).sh"
bash Miniforge3-$(uname)-$(uname -m).sh
```

2. Restart the terminal for the changes to take effect.

xxx

3. Create and activate the conda environment via miniforge.

```{bash, eval=FALSE}
conda create --name scib python=3.11

conda activate scib
```

4. Install dependencies: `python-annoy` (needed for bbknn and scanorama) and `leidenalg` (needed for scMC).

```{bash, eval=FALSE}
pip install leidenalg numpy python-igraph
conda install -c conda-forge python-annoy pooch



```

5. Install the data integration softwares.

```{bash, eval=FALSE}
pip install scvi-tools scanpy scpython bbknn anndata scanorama 
```

6. Record the path to the `scib` environment.
```{bash, eval=FALSE}
conda env list
where.exe python
which python
```


## SCIB package

You can install the latest version scib like so:

```{r, eval=FALSE}
library(remotes)

install_github("mthimma/scib")
```


# LOAD DATA FOR DEMONSTRATION

We load the required packages:

```{r}
library(tidyverse)
library(Seurat)
library(SeuratWrappers)
library(reticulate)
library(scib)

```

path to `scib` conda environment:

```{r}
conda_env <- "C:/Users/aramasamy/AppData/Local/mambaforge/envs/scib"

use_python( file.path(conda_env, "python.exe") )
```

Next, we will load the `pbmc.demo` dataset. This is the same dataset that used in the Seurat vignette and contains 2,700 peripheral blood mononuclear cells (PBMC) from 10X genomics. See `help(pbmc.demo)` for more details.

```{r}
data(pbmc.demo)
pbmc.demo
```


<!-- We also need to specify the conda environment which contain python packages and dependencies for BBKNN, Scanaroma and scVI. -->
<!-- For example, we can do this in Windows via: -->

<!-- ```{r} -->
<!-- conda_env <- conda_list() %>%  -->
<!--   filter(name == "scib") %>%  -->
<!--   pull(python) %>%  -->
<!--   gsub("/python.exe", "", .) -->

<!-- conda_env -->
<!-- ``` -->


## Quality control

Like all single-cell dataset, we need to perform quality control to ensure high quality dataset. Here we conduct a lenient quality control.

```{r}
pbmc.demo <- pbmc.demo[ , !is.na(pbmc.demo$seurat_annotations)]  ## fix

pbmc.demo$percent.mt <- PercentageFeatureSet(pbmc.demo, pattern = "^MT-|^Mt-")

VlnPlot(pbmc.demo, c("nCount_RNA", "nFeature_RNA", "percent.mt"))

pbmc.demo <- subset(pbmc.demo,
                    nFeature_RNA > 200 &
                      nFeature_RNA < 2500 &
                      percent.mt   < 5)
```

Aside: After quality control

```{r}
pbmc.demo

VlnPlot(pbmc.demo, c("nCount_RNA", "nFeature_RNA", "percent.mt"))
```

## Ground truth

This dataset includes `seurat_annotations` provided by the Seurat co-authors. We also included a `batch` column which is artificially created for this package demonstration.

```{r}
pbmc.demo@meta.data %>%
  glimpse()
```

```{r}
pbmc.demo@meta.data %>% 
  janitor::tabyl(seurat_annotations, batch)
```

# DATA INTEGRATION

Aside: Seurat v5 stores information in the following layers

* `counts` layer: raw un-normalized counts
* `data` layer: normalized data, which is set correctly after `NormalizeData()`
* `scale.data` layer: z-scored/variance-stabilized data, which is set after `ScaleData()`

Here are the layers before the split

```{r}
Layers(pbmc.demo)
```


## Step 1: Split the Seurat object

We will split the RNA assay by the grouping factor that defines the batch effect. In reality, batch effects may arise from differences between donors, single-cell technology, sequencing platforms, timing, reagents, or experimental conditions across laboratories.

```{r}
pbmc.demo[["RNA"]] <- split(pbmc.demo[["RNA"]], f = pbmc.demo$batch)
```

Aside: We now can see that the counts and data layer has been split into three layers each.

```{r}
Layers(pbmc.demo)
```

## Step 2: Preprocess each split

Next we need to run the standard preprocessing steps for each split. This can be done via:

```{r}
pbmc.demo <- pbmc.demo %>%
  NormalizeData(verbose = FALSE) %>%
  FindVariableFeatures(verbose = FALSE) %>%
  ScaleData(verbose = FALSE) %>%
  RunPCA(verbose = FALSE)
```

We can see this has created the `pca` reduction based on the unintegrated data. We can determine the dimensional of this dataset from the Elbow plot to be approximately 10.

```{r}
Reductions(pbmc.demo)

ElbowPlot(pbmc.demo, ndims = 50, reduction = "pca")

ndims <- 10
```


## Step 3: Data integration step

This is the workhorse of the data integration process. Here is an example on how to execute this with Harmony. The integrated embedding is stored in the `integrated.harmony` reduction.

```{r, echo=FALSE}
set.seed(123)  # to fix the random seed for consistency in the .md output
```


```{r, warning=FALSE, message=FALSE}
pbmc.demo <- IntegrateLayers2(pbmc.demo,
                              method         = "HarmonyIntegration",
                              orig.reduction = "pca",
                              new.reduction  = "integrated.harmony",
                              verbose        = FALSE)

Reductions(pbmc.demo)
```

Other available methods that can be used with `IntegrateLayers()`:

| method$^*$ | Package for wrapper | Publication                                     | DOI                                           |
|---         |---                  |---                                              |---                                            |
| CCA        | Seurat              | Stuart T. et al. Cell (2019)                    | https:://doi.org/10.1016/j.cell.2019.05.031   |
| RPCA       | Seurat              | Stuart T. et al. Cell (2019)                    | https:://doi.org/10.1016/j.cell.2019.05.031   |
| Harmony    | Seurat              | Korsunsky, I. et al. Nat Methods (2019)         | https://doi.org/10.1038/s41592-019-0619-0     |
| FastMNN    | SeuratWrapper       | Haghverdi, L. et al. Nat. Biotechnol. (2018)    | https://doi.org/10.1038/s41421-019-0114-x     |
| SCVI       | SeuratWrapper       | Lopez, R.et al. Nat Methods (2018)              | https://doi.org/10.1038/s41592-018-0229-2     |
| BBKNN      | scib (this package) | Krzysztof Polański et al. Bioinformatics (2020) | https://doi.org/10.1093/bioinformatics/btz625 |
| CONOS      | scib (this package) | Barkas, N et al. Nat Methods (2019)             | https://doi.org/10.1038/s41592-019-0466-z     |
| LIGER      | scib (this package) | Liu, J.et al. Nat. Protocols (2020)             | https://doi.org/10.1038/s41596-020-0391-8     |
| SCANORAMA  | scib (this package) | Hie, B et al. Nat Biotechnology (2019)          | https://doi.org/10.1038/s41587-019-0113-3     |
| SCMC       | scib (this package) | Zhang, L. et al. Genome Biology (2021)          | https://doi.org/10.1186/s13059-020-02238-2    |


$^*$ The specific method can be called by appending the word "Integration" to the method name (e.g. HarmonyIntegration).


## Step 4: Processing the integrated data

We can now run the non-linear dimension reduction on the integrated embedding and save it. Here the `ndims` is the dimensionality of the dataset that we determined earlier.

```{r, warning=FALSE}
pbmc.demo <- RunUMAP(pbmc.demo,
                     dims           = 1:ndims,
                     reduction      = "integrated.harmony",
                     reduction.name = "umap.harmony",
                     verbose        = FALSE
)
```

Next we cluster the cells at certain resolution based on the integrated embedding. We chose `res = 0.80` here but you should determine this for your own dataset.

```{r}
pbmc.demo <- FindNeighbors(pbmc.demo, reduction = "integrated.harmony")

pbmc.demo <- FindClusters (pbmc.demo, res = 0.80, cluster.name = "cluster.harmony")

DimPlot(pbmc.demo,
        reduction = "umap.harmony",
        group.by  = "cluster.harmony",
        label     = TRUE) +
  NoLegend()
```


## (optional) Step 5: Cleanup after data integration

After the data integration, we no longer need the counts and data to be split by the batch variable. 
Similarly, we do not need the `scale.data` layer which being a non-sparse matrix takes a lot of space.

```{r}
Layers(pbmc.demo)

object.size(pbmc.demo) %>% format(units = "Mb")
```

We can join the layers and remove the `scale.data` to reduce object size.

```{r}
pbmc.demo <- JoinLayers(pbmc.demo)
pbmc.demo[["RNA"]]$scale.data  <- NULL
```

We can confirm this by checking the layers and observing the reduction in object size (from 100Mb to 60Mb).

```{r}
Layers(pbmc.demo)  
object.size(pbmc.demo) %>% format(units = "Mb")  
```


# EVALUATION

## Cross tabulation with the ground truth

We can compare the clusters based on the integrated embedding vs. the ground truth provided by the Seurat authors:

```{r, eval = FALSE}
tb <- table(pbmc.demo$seurat_annotations, pbmc.demo$cluster.harmony)

print.table(tb, zero.print = ".")
#                0   1   2   3   4   5   6   7   8   9  10  11
# Naive CD4 T  382  22   . 239   .  54   .   .   .   .   .   .
# Memory CD4 T   9 339   .  88   .  42   .   .   .   5   .   .
# CD14+ Mono     .   .   .   . 251   . 224   2   .   .   3   .
# B              .   1 343   .   .   .   .   .   .   .   .   .
# CD8 T          .   5   1   .   . 130   .   .   5 130   .   .
# FCGR3A+ Mono   .   .   .   .   5   .   1 156   .   .   .   .
# NK             .   .   .   .   .   .   .   . 148   7   .   .
# DC             .   .   .   .   .   .   .   .   .   .  32   .
# Platelet       .   .   .   .   .   .   .   .   .   .   .  14
```

There are several ways of estimating the performance of the data integration technique. Accuracy is one of the simpler method. 

1. Assign every cluster to one label using simple majority 

```{r}
pbmc.demo$ann.harmony <- case_match(pbmc.demo$cluster.harmony,
                                    "0" ~ "Naive CD4 T",
                                    "1" ~ "Memory CD4 T",
                                    "2" ~ "B",
                                    "3" ~ "Naive CD4 T",
                                    "4" ~ "CD14+ Mono",
                                    "5" ~ "CD8 T",
                                    "6" ~ "CD14+ Mono",
                                    "7" ~ "FCGR3A+ Mono",
                                    "8" ~ "NK",
                                    "9" ~ "CD8 T",
                                    "10" ~ "DC",
                                    "11" ~ "Platelet")
```
2. Add up the numbers that correspond to this majority based assignment: 382 + 339 + 343 + 239 + 251 + 130 + 224 + 156 + 148 + 130 + 32 + 14 which adds up to 2,388.

3. Divide by the total number of cells. i.e. 2388 / 2638 which gives 90.5%.

```{r}
mean(pbmc.demo$ann.harmony == pbmc.demo$seurat_annotations)
```


However, there are more sophisticated metrics for evaluation which we cover next.


## Evaluate the data integration

```{r, eval = FALSE}
results <- run_eval_metrics(pbmc.demo, 
                            reduction      = "integrated.harmony", 
                            predicted.cn   = "cluster.harmony",
                            groundtruth.cn = "seurat_annotations")

data.frame(results)
#                            results
# n_celltype            9.0000000000
# ari_celltype          0.6097111530
# nmi_celltype          0.7054868932
# asw_celltype          0.1027220944
# lisi_mean_celltype    0.1459063973
# lisi_median_celltype  0.0674748762
# n_batch               3.0000000000
# ari_batch            -0.0002237677
# nmi_batch             0.0012540544
# asw_batch            -0.0028503476
# lisi_mean_batch       0.7411130657
# lisi_median_batch     0.7707188999
# kbet_batch            0.9753787879
```


# EVALUATING MULTIPLE METHODS

## Data preparation

Let us demonstrate how to run and evaluate mutiple data integration techniques. Let us reset the pbmc.demo data first.

```{r}
rm(pbmc.demo)
data(pbmc.demo, package = "scib")
pbmc.demo <- pbmc.demo[ , !is.na(pbmc.demo$seurat_annotations)]  ## fix
```

Next we can run through Steps 1 - 2 to prepare the data again
```{r}
pbmc.demo[["RNA"]] <- split(pbmc.demo[["RNA"]], f = pbmc.demo$batch)

pbmc.demo <- pbmc.demo %>%
  NormalizeData(verbose = FALSE) %>%
  FindVariableFeatures(verbose = FALSE) %>%
  ScaleData(verbose = FALSE) %>%
  RunPCA(verbose = FALSE)

ndims <- 10  # determined from Elbowplot(pbmc.demo) earlier
```


## Multi method evaluation

Let's demonstrate Steps 3 - 4 with all the methods.

```{r, warnings=FALSE, message=FALSE}
methods <- c(
  "BBKNNIntegration", 
  "CCAIntegration", 
  "CONOSIntegration", 
  "FastMNNIntegration", 
  "HarmonyIntegration", 
  "LIGERIntegration",
  "RPCAIntegration", 
  "SCANORAMAIntegration",
  "SCMCIntegration", 
  "scVIIntegration")

big.res <- list()

for(method in methods){

  cat(method, "\t", date(), "\n")
  method.name <- gsub("Integration", "", method) %>% str_to_lower()
  method.name
  
  ## Step 3
  pbmc.demo <- IntegrateLayers2(object         = pbmc.demo,
                                method         = method,
                                orig.reduction = "pca",
                                new.reduction  = paste0("integrated.", method.name),
                                verbose        = FALSE,
                                ndims          = ndims,
                                resolution     = 0.80,
                                conda_env      = conda_env
  )
  
  ## Step 4
  pbmc.demo <- FindNeighbors(pbmc.demo, reduction = paste0("integrated.", method.name))
  pbmc.demo <- FindClusters (pbmc.demo, res = 0.80, cluster.name = paste0("cluster.", method.name))
  pbmc.demo$seurat_clusters <- NULL
  
  ## Evaluation step
  res <- run_eval_metrics(pbmc.demo, 
                          reduction      = paste0("integrated.", method.name),  
                          predicted.cn   = paste0("cluster.", method.name),
                          groundtruth.cn = "seurat_annotations")
  
  res <- data.frame(res)
  colnames(res) <- method.name
  big.res[[method]]  <- res
  
  ## Cleanup
  pbmc.demo@graphs$RNA_nn  <- NULL
  pbmc.demo@graphs$RNA_snn <- NULL
  pbmc.demo@meta.data[ , paste0("cluster.", method.name)] <- NULL
  rm(res)
}

big.res <- do.call("cbind", big.res)

round(big.res, digits = 2)
```

|                     |  bbknn|    cca| conos| fastmnn| harmony| liger|   rpca| scanorama|   scmc|   scvi|
|:--------------------|------:|------:|-----:|-------:|-------:|-----:|------:|---------:|------:|------:|
|n_celltype           |  9.000|  9.000| 9.000|   9.000|   9.000| 9.000|  9.000|     9.000|  9.000|  9.000|
|ari_celltype         |  0.000|  0.662| 0.327|   0.630|   0.643| 0.489|  0.670|     0.642|  0.688|  0.000|
|nmi_celltype         |  0.005|  0.744| 0.519|   0.748|   0.735| 0.584|  0.771|     0.725|  0.744|  0.006|
|asw_celltype         | -0.051|  0.125| 0.345|   0.133|   0.101| 0.224|  0.107|     0.081|  0.200| -0.107|
|lisi_mean_celltype   |  0.571|  0.141| 0.133|   0.125|   0.139| 0.123|  0.159|     0.087|  0.099|  0.511|
|lisi_median_celltype |  0.571|  0.059| 0.045|   0.081|   0.065| 0.051|  0.078|     0.059|  0.055|  0.515|
|n_batch              |  3.000|  3.000| 3.000|   3.000|   3.000| 3.000|  3.000|     3.000|  3.000|  3.000|
|ari_batch            |  0.000|  0.000| 0.139|   0.000|   0.000| 0.171|  0.000|     0.000|  0.000|  0.000|
|nmi_batch            |  0.001|  0.001| 0.217|   0.001|   0.001| 0.234|  0.001|     0.001|  0.001|  0.001|
|asw_batch            | -0.005| -0.026| 0.019|  -0.015|  -0.003| 0.092| -0.006|    -0.006| -0.007| -0.012|
|lisi_mean_batch      |  0.868|  0.458| 0.225|   0.539|   0.739| 0.167|  0.748|     0.740|  0.782|  0.784|
|lisi_median_batch    |  0.889|  0.454| 0.031|   0.514|   0.770| 0.049|  0.777|     0.768|  0.810|  0.820|
|kbet_batch           |  0.981|  0.100| 0.131|   0.235|   0.964| 0.049|  0.771|     0.911|  0.947|  0.972|

